---
title: "CLK reproduction"
output:
  html_document:
    df_print: paged
---

## Introduction
CLK-dependent exon recognition and conjoined gene formation revealed with a novel small molecule inhibitor

[CLK](https://www.ncbi.nlm.nih.gov/gene?Db=gene&Cmd=ShowDetailView&TermToSearch=1195) encodes a member of the CDC2-like (or LAMMER) family of dual specificity protein kinases. In the cell nucleus, the encoded protein phosphorylates serine/arginine-rich proteins involved in pre-mRNA processing, releasing them into the nucleoplasm. The choice of splice sites during pre-mRNA processing may be regulated by the concentration of transacting factors, including serine/arginine-rich proteins. Therefore, the encoded protein may play an indirect role in governing splice site selection.

The authors describe a new compound, T3, a CLK small molecule inhibitor which shows superior potentcy and selectivity than the current standard, KH-CB19.

Of particular interest in the formation of conjoined genes (CG) when CLK is inhibited.


### Sequencing

```{r}
library(knitr)
library(ggplot2)
library(dplyr)
library(stringr)
library(biomaRt)
library(Gviz)
metadata<-read.csv("../metadata.csv")
```

`nrow(metadata)` were obtained

### Breakdowns
#### By platform
```{r}
metadata %>% group_by(Platform) %>% summarize(samples=n(),median_reads=median(spots))
```

#### By cell line

HCT116 is a human colorectal carcinoma (i.e. malignant or "transformed") cell line. 184-hTERT-L2 cell line is derived from human mammary epithelial cells immortalized by transduction with hTERT.

```{r}
metadata$cell_line<-as.vector(str_extract_all(metadata$SampleName,'(HCT116|184-hTert)',simplify=TRUE))
metadata %>% group_by(cell_line) %>% summarize(samples=n()) %>% arrange(-samples)
```

#### By treatment

Various silencing transfections were tested on CLK itself, splicing factors such as U2AF2 and SRSF9, TIA1 and CPEB RNA-binding proteins
siCLK, siCPEB ,siDAZA, siHNRNPH, siKHDRBS1, siLIN28A, siELAVL1, siHNRNPC, siHNRNPF, siU2AF2, siTIA, siSRSF, siSRRM, siSFPQ, siSAMD4B, 
RNA recognition motif (RRMs) and splicing enhancers are also tested.


```{r}
metadata %>% group_by(SampleName) %>% summarize(samples=n()) %>% arrange(-samples)
```

### Alignment and rMATS-ISO
Alignment was performed with STAR against hg38, using the preferred iGenomes gtf reference. Both "stranded" and "unstranded" RNA-Seq samples were combined.

```
STAR --runMode alignReads \
     --outSAMtype BAM SortedByCoordinate \
     --limitBAMsortRAM ${bytes} \
     --readFilesCommand zcat \
     --outFilterType BySJout   --outFilterMultimapNmax 20 \
     --outFilterMismatchNmax 999   --alignIntronMin 25  \
     --alignIntronMax 1000000   --alignMatesGapMax 1000000 \
     --alignSJoverhangMin 8   --alignSJDBoverhangMin 5 \
     --sjdbGTFfile GRCh38_star/genes.gtf \
     --genomeDir GRCh38_star \
     --runThreadN ${cpus} \
     --outFileNamePrefix ${sample}.  \
     --readFilesIn  ${sample}_1.fastq.gz ${sample}_2.fastq.gz
```

RMATS-ISO was run using the gencode.v28.annotation.gtf, roughly equivalent to the Ensembl gtf.


### RMATS-EM output

The following columns are returned from rMATS-EM:
```
 [1] "ASM_name"                        "total_isoforms"                  "total_exons"                     "total_read_count_group_1"        "total_read_count_group_2"       
 [6] "p_value"                         "test_statistic"                  "isoform_inclusion_group_1"       "isoform_inclusion_group_2"       "isoform_inclusion_constrained"  
[11] "variance_group_1"                "variance_group_2"                "variance_constrained"            "dirichlet_parameter_group_1"     "dirichlet_parameter_group_2"    
[16] "dirichlet_parameter_constrained" "paired_isoform_pvalues"          "isoform_index"                   "merge_info"   
```

### Dose dependent splicing patterns of T3 in HCT116

```{r hct116}
pthresh<-0.01
doses<-c('0.05','0.1','0.5','1.0','5.0')
control<-"untreated"
```

```{r plotdoseresponse}
em_all<-NULL
for(dose in doses){
  em<-read.table(paste0(control,'_vs_',dose,'/EM_out/EM.out'),comment.char = '',strip.white = TRUE, header=TRUE)
  em$dose<-paste0(control,'_vs_',dose)
  if(!is.null(em_all)){
    em_all<-rbind(em_all,em)
  }else{
    em_all<-em
  }
}

#set p_value of 0 to a token dummy minimum, display only highly significant assemblies
#ggplot(em_all %>% filter(!is.na(p_value)) %>% filter(p_value<=0.01) %>% rowwise() %>% dplyr::mutate(p_value = max(p_value,1e-16)),aes(-log(p_value)))+geom_histogram(binwidth=1)+facet_wrap(.~dose)
ggplot(em_all %>% filter(!is.na(p_value)) %>% rowwise() %>% mutate(sig=ifelse(p_value<=pthresh,paste0("p<=",pthresh),paste0("p>",pthresh))),aes(sig))+geom_bar(stat="count")+facet_wrap(.~dose)
```


```{r tabledoseresponse, warning=FALSE}
em_all %>% filter(!is.na(p_value)) %>% rowwise() %>% mutate(sig=p_value<=pthresh) %>% group_by(dose,sig) %>% summarize(cnt=n()) %>% reshape2::acast(dose~sig) -> sig_table
dimnames(sig_table)[[2]]<-c(paste0("p>",pthresh),paste0("p<=",pthresh))
sig_table<-cbind(sig_table,total=rowSums(sig_table))
sig_table<-cbind(sig_table,sig_frac=round(sig_table[,2]/sig_table[,3],2))
fold_change<-c(0,sapply(1:(nrow(sig_table)-1),function(x){(sig_table[x+1,2]/sig_table[x+1,3])/(sig_table[x,2]/sig_table[x,3])}))
sig_table<-cbind(sig_table,fold_change=fold_change)
knitr::kable(sig_table)
```

The number of assemblies in which significant AS events (p<`r pthresh`) were observed as a fraction of all assemblies was highest in the `untreated_vs_1.0` group, although the greatest fold increase occurs at 0.5uM `r max(fold_change)` reported in the paper as 4.1.

### Dose dependent splicing patterns of T3 in 184-hTert
```{r htert184}
doses<-c('0.5-184','1.0-184','5.0-184')
control<-"untreated184"
```

```{r htert184plot, ref.label='plotdoseresponse'}
```

```{r htert184table, ref.label='tabledoseresponse'}
```

### Clusters of interest in T3 0.5
```{r doi}
dose<-c('0.5')
control<-"untreated"
em<-read.table(paste0(control,'_vs_',dose,'/EM_out/EM.out'),comment.char = '',strip.white = TRUE, header=TRUE)
coor<-read.table(paste0(control,'_vs_',dose,'/ISO_classify/ISO_module_coor.txt'),comment.char = '',strip.white = TRUE, header=FALSE,sep=":",col.names = c("chr","strand","start","end"))
gene<-read.table(paste0(control,'_vs_',dose,'/ISO_classify/ISO_module_gene.txt'),comment.char = '',strip.white = TRUE, header=FALSE,sep="_",col.names= c("asm","hugo","ensg"))
#type<-read.table(paste0(control,'_vs_',dose,'/ISO_classify/ISO_module_type.txt'),comment.char = '',strip.white = TRUE, header=FALSE)
#typesummary<-
```

#### Top hit: ASM3356

In teh 
```{r}
knitr::kable(em %>% filter(test_statistic==max(em$test_statistic,na.rm = TRUE)))
gene %>% filter(asm=='ASM3356')
coords<-coor[which(em$ASM_name=='ASM#3356'),]
```

```{r}
dose="0.5"
#treated
metadata %>% filter(str_detect(SampleName,dose)) %>% filter(str_detect(SampleName,'T3')) %>% filter(Platform=='ILLUMINA') %>% filter(str_detect(SampleName,'HCT116'))  %>% dplyr::pull(Run) -> treated
metadata %>% filter(str_detect(SampleName,'Untreated')) %>% filter(Platform=='ILLUMINA') %>% filter(str_detect(SampleName,'HCT116')) %>% dplyr::pull(Run) -> untreated

#aws s3 cp s3://panorama-clk-repro/SRP091981/SRR5009487.Aligned.sortedByCoord.out.bam SRP091981/
atreated<-"SRR5009487"
auntreated<-"SRR5009474"
afrom <- coords$start - 10000
ato <- coords$end + 10000
chr <- as.character(coords$chr)
treatedTrack <- AlignmentsTrack(paste0("../SRP091981/",atreated,".Aligned.sortedByCoord.out.bam"), isPaired = TRUE,name = metadata %>% filter(Run==atreated) %>% pull(SampleName) %>% as.character())
untreatedTrack  <- AlignmentsTrack(paste0("../SRP091981/",auntreated,".Aligned.sortedByCoord.out.bam"), isPaired = TRUE, name = metadata %>% filter(Run==auntreated) %>% pull(SampleName) %>% as.character())
options(ucscChromosomeNames=TRUE)

bmt <- BiomartGeneRegionTrack(genome = "hg38", name="ENSEMBL", chromosome = chr, start = afrom, end = ato,biomart=biomaRt::useMart(biomart="ensembl",dataset="hsapiens_gene_ensembl") , stacking = "full")

plotTracks(c(bmt,untreatedTrack,treatedTrack), from = afrom, to = ato, chromosome = chr, type = c("coverage","sashimi"))
```